---
title: "Understanding World Population Dynamics"
subtitle: "Assignment 1 - PSYC593"
format:
  html: default
    self-contained: true
project:
  output-dir: ./doc/
date: 9-8-2023
author: Belgin Unal
---

<!-- These problems come from Imai, K. & Webb Williams, N. (2018). Quantitative Social Science: An introduction -->

<!-- in tidyverse. Princeton University Press. -->

<!-- The code giving partial solutions by Víctor H Cervantes -->

Understanding population dynamics is important for many areas of social
science. We will calculate some basic demographic quantities of births
and deaths for the world's population from two time periods: 1950 to
1955 and 2005 to 2010. We will analyze the following CSV data files -
`Kenya.csv`, `Sweden.csv`, and `World.csv`. Each file contains
population data for Kenya, Sweden, and the world, respectively. The
table below presents the names and descriptions of the variables in each
data set.

<!-- Note the syntax to create a table in markdown -->

<!-- Before the table there must be an empty line, and -->

<!-- the table begins with a line of dashes. -->

<!-- The column names are separated with spaces -->

<!-- and the first character must be a space -->

<!-- Then, segments of repeated dashes indicate the column separations (note -->

<!-- the space between groups of dashes before the character D of the column Description) -->

| Name       | Description                                                                             |
|:------------------|:---------------------------------------------------|
| `country`  | Abbreviated country name                                                                |
| `period`   | Period during which data are collected                                                  |
| `age`      | Age group                                                                               |
| `births`   | Number of births in thousands (i.e., number of children born to women of the age group) |
| `deaths`   | Number of deaths in thousands                                                           |
| `py.men`   | Person-years for men in thousands                                                       |
| `py.women` | Person-years for women in thousands                                                     |

Source: United Nations, Department of Economic and Social Affairs,
Population Division (2013). *World Population Prospects: The 2012
Revision, DVD Edition.*

```{r load-packages, message = FALSE, warning = FALSE}
# Load packages ----
library(tidyverse)
library(rprojroot)
library(here)
```

Identify the subdirectories, these will be useful when we need to read
the data - and if we need to derive some results from the data. (Using
either `rprojroot` or `here` entirely avoids the issues associated
setting the working directory.)

```{r path-variables}
here_path <- here::here()     

# Some of the subdirectories
code_path <- file.path(here_path, "qmd")  
docs_path <- file.path(here_path,"doc")
data_path <- file.path(here_path, "data")
# results_path <- file.path(crit_path, "results") # in case we create some figures and tables
```


```{r read-data, message = FALSE}
# Read data ----
world_data  <- readr::read_csv(file = file.path(data_path, "World.csv"))
kenya_data  <- readr::read_csv(file = file.path(data_path, "Kenya.csv"))
sweden_data <- readr::read_csv(file = file.path(data_path, "Sweden.csv"))
```

<!-- The `read_csv` function from the package `readr` -->

<!-- is part of the `tidyverse`. It will read the data into a -->

<!-- special `data.frame` with more structure known as a `tibble` -->

The data are collected for a period of 5 years where *person-year* is a
measure of the time contribution of each person during the period. For
example, a person that lives through the entire 5 year period
contributes 5 person-years whereas someone who only lives through the
first half of the period contributes 2.5 person-years. Before you begin
this exercise, it would be a good idea to directly inspect each data
set. In R, this can be done with the `View` function, which takes as its
argument the name of a `data.frame` to be examined. Alternatively, in
RStudio, double-clicking a `data.frame` in the `Environment` tab will
enable you to view the data in a spreadsheet-like view.

## Question 1

We begin by computing *crude birth rate* (CBR) for a given period. The
CBR is defined as: $$ 
    \text{CBR} 
     =  \frac{\text{number of births}}{\text{number of person-years lived}}
  $$
<!-- It is also possible to include $\LaTeX$ syntax directly in the document. -->

Compute the CBR for each period, separately for Kenya, Sweden, and the
world. Start by computing the total person-years, recorded as a new
variable within each existing `data.frame` via the `$` operator, by
summing the person-years for men and women. Then, store the results as a
vector of length 2 (CBRs for two periods) for each region with
appropriate labels. You may wish to create your own function for the
purpose of efficient programming. Briefly describe patterns you observe
in the resulting CBRs.

### Answer 1

```{r compute-person-years}
# Create new variable py = total person years for each data set
# py is for person-years
world_data$py  <- world_data$py.men + world_data$py.women
kenya_data$py  <- kenya_data$py.men + kenya_data$py.women
sweden_data$py <- sweden_data$py.men + sweden_data$py.women
```

<!-- Whenever you find that you will repeat the same calculation -->

<!-- on different sets of data, **consider** creating a function -->

<!-- instead of repeating the code for each piece of data. -->

```{r compute-cbr}
# Function to compute the Crude Birth Rate (CBR)
# The function takes the data frame as the input, and computes the CDR
compute_cbr <- function(population_data) {
  population_data %>% 
    group_by(period) %>% 
      summarise(cbr = sum(births) / sum(py)) %>% 
        pull()
}
```

<!-- The "pipe" `%>%` from package `magrittr` takes the output -->

<!-- from the left hand side and makes it the first argument of -->

<!-- the expression on the right hand side. -->

<!-- Starting in R 4.1.0, there is a base pipe operator `|>` -->

<!-- which has a similar behavior. -->

```{r apply-cbr}
# Compute the CBR for each data set
(worldcbr  <- compute_cbr(world_data))
(kenyacbr  <- compute_cbr(kenya_data))
(swedencbr <- compute_cbr(sweden_data))
```

Answer: According to these data, it looks like the Crude Birth Rate
(CBR) is lower in Sweden and higher in Kenya. CBR provides insights into
fertility trends, indicating that birth rates in Sweden are below the
global average, while in Kenya, they are higher.We can also see a decreasing
trend in both countries, as well as in the world, over time. Which means that 
people seem to be having fewer children over time, and this pattern is also true
for Kenya, where the fertility is higher.

## Question 2

The CBR is easy to understand but contains both men and women of all
ages in the denominator. We next calculate the *total fertility rate*
(TFR). Unlike the CBR, the TFR adjusts for age compositions in the
female population. To do this, we need to first calculate the *age
specific fertility rate* (ASFR), which represents the fertility rate for
women of the reproductive age range $[15, 50)$. The ASFR for age range
$[x, x+\delta)$, where $x$ is the starting age and $\delta$ is the width
of the age range (measured in years), is defined as: $$
    \text{ASFR}_{[x,\ x+\delta)} 
    \ = \ \frac{\text{number of births to women of age $[x,\ x+\delta)$}}{\text{Number of person-years lived by women of age $[x,\ x+\delta)$}}
  $$ Note that square brackets, $[$ and $]$, include the limit whereas
parentheses, $($ and $)$, exclude it. For example, $[20, 25)$ represents
the age range that is greater than or equal to 20 years old and less
than 25 years old. In typical demographic data, the age range $\delta$
is set to 5 years. Compute the ASFR for Sweden and Kenya as well as the
entire world for each of the two periods. Store the resulting ASFRs
separately for each region. What does the pattern of these ASFRs say
about reproduction among women in Sweden and Kenya?

### Answer 2

```{r compute-asfr}
# Function to compute Age specific fertility rate (ASFR)
# This function also takes the data frame of the country as an input, computes the ASFR values, and add that as a column to the data frame
compute_asfr <- function(population_data) {
  population_data %>% 
    mutate(asfr = births / py.women)
}
```

```{r apply-asfr}
# Compute ASFR for each data set
world_data  <- compute_asfr(world_data)
kenya_data  <- compute_asfr(kenya_data)
sweden_data <- compute_asfr(sweden_data)
```

```{r compare-asfr}
# Compare ASFRs for Kenya and Sweden
kenya_data$asfr
sweden_data$asfr
```

Answer: The data align with our observations of the Crude Birth Rate
(CBR) values. In both Kenya and Sweden, the Age-Specific Fertility Rate
(ASFR) values rise significantly from ages 20-24 to 40-44, which is
understandable as these are the typical fertile years for women. This
increase, however, much more pronounced in Kenya, indicating that women in Kenya
are more likely to have children as compared to women in Sweden.

## Question 3

Using the ASFR, we can define the TFR as the average number of children
women give birth to if they live through their entire reproductive age.
$$
  \text{TFR} 
   =   \text{ASFR}_{[15,\ 20)} \times 5 + \text{ASFR}_{[20,\ 25)} \times 5 
  + \dots + \text{ASFR}_{[45,\ 50)} \times 5
  $$

We multiply each age-specific fertility rate rate by 5 because the age
range is 5 years. Compute the TFR for Sweden and Kenya as well as the
entire world for each of the two periods. As in the previous question,
continue to assume that women's reproductive age range is $[15, 50)$.
Store the resulting two TFRs for each country or the world as a vector
of length two. In general, how has the number of women changed in the
world from 1950 to 2000? What about the total number of births in the
world?

### Answer 3

```{r compute-tfr}
# Function to compute the total fertility rate (TFR)
# This function takes the country's data frame as input, and computes ASFR values
compute_tfr <- function(population_data) {
  population_data %>% 
    group_by(period) %>% 
      summarise(tfr=5 * sum(asfr)) %>% 
        pull()
}
```



```{r apply-tfr}
# Compute the TFR for each data set
(world_tfr  <- compute_tfr(world_data))
(kenya_tfr  <- compute_tfr(kenya_data))
(sweden_tfr <- compute_tfr(sweden_data))
```

```{r compare-tfr}
# Compute totals of women and births in the world by period
(
  world_data %>% 
    group_by(period) %>% 
    summarise(total_women = sum(py.women),
              total_births = sum(births)) -> 
    totals_world
)

# Compare how much these totals have changed
(changes_totals <- totals_world[2,-1] / totals_world[1,-1])
```

Answer: Of course, between the periods of 1950-1955 and 2005-2010, there
has been a rise in both the population of women and the total number of
births. However, the increase in the number of women has outpaced the
increase in the number of births. This data indicates a trend of women
having fewer children over time.
When we look at the data from Kenya and Sweden, even though the values from
Kenya are higher (indicating a greater fertility rate), both countries show a
descresing trend in fertility over time.

## Question 4

Next, we will examine another important demographic process: death.
Compute the *crude death rate* (CDR), which is a concept analogous to
the CBR, for each period and separately for each region. Store the
resulting CDRs for each country and the world as a vector of length two.
The CDR is defined as: $$ 
    \text{CDR} 
     =  \frac{\text{number of deaths}}{\text{number of person-years lived}}
  $$ Briefly describe patterns you observe in the resulting CDRs.

```{r compute-cdr}
# Function to compute the Crude death rate (CDR)
# This function takes the country's data frame as input, and computes the CDR values
compute_cdr <- function(population_data) {
  population_data %>% 
    group_by(period) %>% 
      summarise(cbr = sum(deaths) / sum(py)) %>% 
        pull()
}
```

```{r apply-cdr}
# Compute the CDR for each data set
(world_cdr  <- compute_cdr(world_data))
(kenya_cdr  <- compute_cdr(kenya_data))
(sweden_cdr <- compute_cdr(sweden_data))
```

Answer: Similar to the lower birth rates, Sweden also has lower death
rates -with a lower Crude Death Rate - compared to Kenya and the global
average. Kenya, on the other hand, is still above the global average -
indicating a higher mortality rate in Kenya. We also see a decline in
the mortality rates over time, to the point that the values we observe
for Kenya and Sweden during the period of 2005-2010 are relatively
similar.

## Question 5

One puzzling finding from the previous question is that the CDR for
Kenya during the period of 2005-2010 is about the same level as that for
Sweden. We would expect people in developed countries like Sweden to
have a lower death rate than those in developing countries like Kenya.
While it is simple and easy to understand, the CDR does not take into
account the age composition of a population. We therefore compute the
*age specific death rate* (ASDR). The ASDR for age range $[x, x+\delta)$
is defined as: $$
    \text{ASDR}_{[x,\ x+\delta)} 
    \ = \ \frac{\text{number of deaths for people of age $[x,\ x+\delta)$}}
               {\text{number of person-years of people of age $[x,\ x+\delta)$}}
  $$ Calculate the ASDR for each age group, separately for Kenya and
Sweden, during the period of 2005-2010. Briefly describe the pattern you
observe.

```{r compute asdr}
# Function to compute Age specific death rate (ASDR)
# This function tajes the country's population data as input and adds a new column
# to the data frame for ASDR values
compute_asdr <- function(population_data) {
  population_data %>% 
    mutate(asdr = deaths / py)
}
```

```{r apply-asdr}
# Compute ASDR for each data set
world_data  <- compute_asdr(world_data)
kenya_data  <- compute_asdr(kenya_data)
sweden_data <- compute_asdr(sweden_data)
```

Answer: When we actually look at the Age-Specific Death Rate (ASDR)
data, we see that in Sweden, a higher number of individuals pass away
after reaching the age of 80+. This discrepancy likely arises from the
fact that Sweden's population skews older compared to Kenya's. This
shows that the previous similarity we observed in CDR rates between
Sweden and Kenya occurs because of the age composition of the
population.

## Question 6

One way to understand the difference in the CDR between Kenya and Sweden
is to compute the counterfactual CDR for Kenya using Sweden's population
distribution (or vice versa). This can be done by applying the following
alternative formula for the CDR. $$
    \text{CDR}
    \ = \ \text{ASDR}_{[0, 5)} \times P_{[0,5)} + \text{ASDR}_{[5, 10)}
    \times P_{[5, 10)} + \cdots 
  $$ where $P_{[x, x+\delta)}$ is the proportion of the population in
the age range $[x, x+\delta)$. We compute this as the ratio of
person-years in that age range relative to the total person-years across
all age ranges. To conduct this counterfactual analysis, we use
$\text{ASDR}_{[x,x+\delta)}$ from Kenya and $P_{[x,x+\delta)}$ from
Sweden during the period of 2005--2010. That is, first calculate the
age-specific population proportions for Sweden and then use them to
compute the counterfactual CDR for Kenya. How does this counterfactual
CDR compare with the original CDR of Kenya? Briefly interpret the
result.

```{r compute-population-proportion}
# Function to compute population proportion by period
# This function takes the country's data frame as input and adds a new column to
# the data frame for the proportion of population for counterfactual CDR
compute_pop_prop <- function(population_data) {
  population_data %>% 
    group_by(period) %>%
    mutate(pop_prop = py / sum(py)) %>%
    ungroup()
}
```

```{r apply-population-proportion}
# Compute population proportion for each data set
world_data  <- compute_pop_prop(world_data)
kenya_data  <- compute_pop_prop(kenya_data)
sweden_data <- compute_pop_prop(sweden_data)
```

```{r compute-counterfact}
# Compute Kenyas CDR Kenya had Sweden's population distribution
mutate(kenya_data,
       temp_cdr = asdr * sweden_data$pop_prop) %>%
  group_by(period) %>% 
  summarise(cdrresweden = sum(temp_cdr))
```

Answer: So the previous CDR results for Kenya were 1950-1955: 0.02396254
and 2005-2010: 0.01038914 There does not seem to be much of a difference
for the period of 1950-1955. However, the counterfactual CDR for
2005-2010 is higher than the original CDR. This does make sense,
considering the older population of Sweden and Kenya's relatively
younger population.
